---
title: "GDA Capstone - Cyclistic"
author: "Edward Yeakel"
date: "2023-07-04"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

<br>

### How did annual members and casual riders use cyclistic bikes differently in Chicago in 2019?

<br>
<br>
First we will install and load necessary packages:
<br>
```{r}
library(tidyverse)
library(lubridate)
```
<br>

## Gathering & Formatting

<br>
<br>
To start, we will load our datasets, one from each quarter in 2019:
<br>
```{r}
q1_trips <- read_csv("~/Desktop/GDA_capstone_cyclistic/Divvy_Trips_2019_Q1.csv")
q2_trips <- read_csv("~/Desktop/GDA_capstone_cyclistic/Divvy_Trips_2019_Q2.csv")
q3_trips <- read_csv("~/Desktop/GDA_capstone_cyclistic/Divvy_Trips_2019_Q3.csv")
q4_trips <- read_csv("~/Desktop/GDA_capstone_cyclistic/Divvy_Trips_2019_Q4.csv")
```
<br>
<br>
Lets take a look at our data and see what we are working with, keeping in mind that our goal at this stage is to combine all four data frames into one.
<br>
```{r}
glimpse(q1_trips)
glimpse(q2_trips)
glimpse(q3_trips)
glimpse(q4_trips)
```
<br>
<br>
Lets rename the columns in each data frame so we can combine them.
<br>
```{r}
q1_2019_trips <- rename(q1_trips,
                        ride_id = trip_id,
                        started_at = start_time,
                        ended_at = end_time,
                        bike_id = bikeid,
                        trip_seconds = tripduration,
                        start_station_id = from_station_id,
                        start_station_name = from_station_name,
                        end_station_id = to_station_id,
                        end_station_name = to_station_name,
                        usertype = usertype,
                        sex = gender,
                        birthyear = birthyear)

q2_2019_trips <- rename(q2_trips,
                        ride_id = "01 - Rental Details Rental ID",
                        started_at = "01 - Rental Details Local Start Time",
                        ended_at = "01 - Rental Details Local End Time",
                        bike_id = "01 - Rental Details Bike ID",
                        trip_seconds = "01 - Rental Details Duration In Seconds Uncapped",
                        start_station_id = "03 - Rental Start Station ID",
                        start_station_name = "03 - Rental Start Station Name",
                        end_station_id = "02 - Rental End Station ID",
                        end_station_name = "02 - Rental End Station Name",
                        usertype = "User Type",
                        sex = "Member Gender",
                        birthyear = "05 - Member Details Member Birthday Year")

q3_2019_trips <- rename(q3_trips,
                        ride_id = trip_id,
                        started_at = start_time,
                        ended_at = end_time,
                        bike_id = bikeid,
                        trip_seconds = tripduration,
                        start_station_id = from_station_id,
                        start_station_name = from_station_name,
                        end_station_id = to_station_id,
                        end_station_name = to_station_name,
                        usertype = usertype,
                        sex = gender,
                        birthyear = birthyear)

q4_2019_trips <- rename(q4_trips,
                        ride_id = trip_id,
                        started_at = start_time,
                        ended_at = end_time,
                        bike_id = bikeid,
                        trip_seconds = tripduration,
                        start_station_id = from_station_id,
                        start_station_name = from_station_name,
                        end_station_id = to_station_id,
                        end_station_name = to_station_name,
                        usertype = usertype,
                        sex = gender,
                        birthyear = birthyear)
```
<br>
<br>
We will now inspect the data frames for consistency:
<br>
```{r}
glimpse(q1_2019_trips)
glimpse(q2_2019_trips)
glimpse(q3_2019_trips)
glimpse(q4_2019_trips)

```
<br>
<br>
The data frames are ready to be combined:
<br>
```{r}
trips_2019 <- bind_rows(q1_2019_trips,q2_2019_trips,q3_2019_trips,q4_2019_trips)
```
<br>
We want to confirm our data frame is correct:
<br>
<br>
```{r}
glimpse(trips_2019)
```
<br>
<br>
Lets create two new columns to help us get a better sense of the data. One column will show trip duration in minutes, the other in hours.
<br>
```{r}
# Find trip duration in minutes and create a column for it
trips_2019['trip_minutes'] <- trips_2019['trip_seconds'] / 60

# Find trip duration in hours and create a column for it
trips_2019['trip_hours'] <- trips_2019['trip_minutes'] / 60
```
<br>
<br>
We will rename the 'usertype' column to fit Cyclistic's current naming conventions.
<br>
```{r}
trips_2019 <- trips_2019 %>%
  mutate(usertype = recode(usertype,
                           "Customer" = "casual",
                           "Subscriber" = "member"))
```
<br>
<br>
Verify usertype column variable name changes:
<br>
```{r}
table(trips_2019$usertype)
```
<br>
<br>
To help with out visualizations later on, lets create columns for day, month, year, month name and weekday:
<br>
```{r}
trips_2019$date <- as.Date(trips_2019$started_at)
trips_2019$month <- format(as.Date(trips_2019$date),"%m")
trips_2019$day <- format(as.Date(trips_2019$date),"%d")
trips_2019$year <- format(as.Date(trips_2019$date),"%Y")
trips_2019$day_of_week <- format(as.Date(trips_2019$date),"%A")
trips_2019$month_name <- format(as.Date(trips_2019$date),"%B")
```
<br>
<br>
We need to put days of the week in order:
<br>
```{r}
trips_2019$day_of_week <- ordered(trips_2019$day_of_week, levels = 
                                       c("Monday", "Tuesday", "Wednesday", 
                                          "Thursday", "Friday", "Saturday",
                                          "Sunday"))
```
<br>
<br>
We also need to put the months of the year in order:
<br>
```{r}
trips_2019$month_name <- ordered(trips_2019$month_name, levels = 
                                              c("January", "February", "March", 
                                                "April", "May", "June", "July",
                                                "August", "September", 
                                                "October", "November", 
                                                "December"))
```
<br>
<br>
Finally, to find the most popular routes, we will create a column for routes by combining the 'start_station_name' column with the 'end_station_name' column:
<br>
```{r}
trips_2019$route <- str_c(trips_2019$start_station_name, 
                                     " to ", trips_2019$end_station_name)
```
<br>
<br>
Lets inspect our new columns:
<br>
```{r}
glimpse(trips_2019)
```
<br>
<br>
<br>

## Cleaning

<br>
<br>
The data does not require much cleaning However, 1,849 rides (0.05%) are longer than 24 hours, to avoid skewed data, we will  remove these, average trip in minutes before removing is 24 minutes, after is 19 minutes.
<br>
```{r}
trips_2019_v2 <- subset(trips_2019,trip_hours < 24)
```
<br>
<br>
<br>

## Analysis

<br>
<br>
We will gather a quick summary based on trip duration, in minutes:
<br>
```{r}
summary(trips_2019_v2$trip_minutes)
```
<br>
<br>
Compare ride lengths of members to casuals:
<br>
```{r}
aggregate(trips_2019_v2$trip_minutes ~ trips_2019_v2$usertype, FUN = mean)
aggregate(trips_2019_v2$trip_minutes ~ trips_2019_v2$usertype, FUN = median)
aggregate(trips_2019_v2$trip_minutes ~ trips_2019_v2$usertype, FUN = min)
aggregate(trips_2019_v2$trip_minutes ~ trips_2019_v2$usertype, FUN = max)
```
<br>
<br>
Find trip average of each usertype by day of the week:
<br>
```{r}
aggregate(trips_2019_v2$trip_minutes ~ 
            trips_2019_v2$day_of_week + trips_2019_v2$usertype, FUN = mean)
```
<br>
<br>
Find trip average of each usertype by month:
<br>
```{r}
aggregate(trips_2019_v2$trip_minutes ~ 
            trips_2019_v2$month_name + trips_2019_v2$usertype, FUN = mean)
```
<br>
<br>
Create a separate data frame with only casual users:
<br>
```{r}
trips_2019_v2_casuals <- subset(trips_2019_v2,usertype == 'casual')
```
<br>
<br>
Create a separate data frame with only members:
<br>
```{r}
trips_2019_v2_members <- subset(trips_2019_v2,usertype == 'member')
```
<br>
<br>
Breeak down of average ride length, by usertype, for each day of the week:
<br>
```{r}
trips_2019_v2 %>% 
  group_by(usertype, day_of_week) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_minutes)) %>%
  arrange(usertype, day_of_week)
```
<br>
<br>
Break down of average ride length, by usertype, for each month of the year:
<br>
```{r}
trips_2019_v2 %>% 
  group_by(usertype, month_name) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_minutes)) %>%
  arrange(usertype, month_name) %>% 
  print(n=24)
```
<br>
<br>
For casuals, where are they most likely to pick up a bike?
<br>
```{r}
trips_2019_v2_casuals %>% 
  group_by(usertype, start_station_name) %>% 
  summarise(Rides = n()) %>% 
  arrange(usertype, desc(Rides))
```
<br>
<br>
For members, where are they most likely to pick up a bike?
<br>
```{r}
trips_2019_v2_members %>% 
  group_by(usertype, start_station_name) %>% 
  summarise(Rides = n()) %>% 
  arrange(desc(Rides))
```
<br>
<br>
For casuals, where are they most likely to drop off a bike?
<br>
```{r}
trips_2019_v2_casuals %>% 
  group_by(usertype, end_station_name) %>% 
  summarise(Rides = n()) %>% 
  arrange(usertype, desc(Rides))
```
<br>
<br>
For members, where are they most likely to drop off a bike?
<br>
```{r}
trips_2019_v2_members %>% 
  group_by(usertype, end_station_name) %>% 
  summarise(Rides = n()) %>% 
  arrange(desc(Rides))
```
<br>
<br>
For casuals, what is the most used bike route?
<br>
```{r}
trips_2019_v2_casuals %>% 
  group_by(usertype, route) %>% 
  summarise(Rides = n()) %>% 
  arrange(desc(Rides))
```
<br>
<br>
For members, what is the most used bike route?
<br>
```{r}
trips_2019_v2_members %>% 
  group_by(usertype, route) %>% 
  summarise(Rides = n()) %>% 
  arrange(desc(Rides))
```
<br>
<br>
For our final analysis, lets look how ridership breaks down by sex:
<br>
```{r}
trips_2019_v2 %>% 
  group_by(usertype, sex) %>% 
  summarise(Rides = n()) %>% 
  arrange(usertype, desc(Rides))
```
<br>
<br>
<br>

## Visualizations

<br>
<br>
Make bar plot comparing rides per month for each usertype:
<br>
```{r}
trips_2019_v2 %>% 
  group_by(month_name, usertype) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_minutes)) %>%
  arrange(usertype, month_name) %>% 
  ggplot(aes(x = month_name, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#50CAF7","#F59669")) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust= 1)) +
  labs(title = "Rides per Month, Memebers vs Casuals ", subtitle = "2019", 
       x = " ", y = "Number of Rides")
```
<br>
<br>
Make bar plot comparing rides per month for casuals:
<br>
```{r}
trips_2019_v2_casuals %>% 
  group_by(month_name, usertype) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_minutes)) %>%
  arrange(usertype, month_name) %>% 
  ggplot(aes(x = month_name, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#50CAF7")) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust= 1)) +
  labs(title = "Rides per Month, Casuals ", subtitle = "2019", 
       x = " ", y = "Number of Rides", fill = " ")
```
<br>
<br>
Create a bar plot to view rides per weekday for each usertype:
<br>
```{r}
trips_2019_v2 %>% 
  group_by(day_of_week, usertype) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_minutes)) %>%
  arrange(usertype, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = usertype))+
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#50CAF7","#F59669")) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Rides per Weekday, Memebers vs Casuals ", subtitle = "2019", 
       x = " ", y = "Number of Rides")
```


Make bar plot comparing rides per weekday for casuals:

```{r}
trips_2019_v2_casuals %>% 
  group_by(day_of_week, usertype) %>%
  summarise(number_of_rides = n(), average_duration = mean(trip_minutes)) %>%
  arrange(usertype, day_of_week) %>% 
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#50CAF7")) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_y_continuous(labels = scales::comma) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust= 1)) +
  labs(title = "Rides per Month, Casuals ", subtitle = "2019", 
       x = " ", y = "Number of Rides", fill = " ")
```
<br>
<br>
Make a bar graph comparing sexes by usertype:
<br>
```{r}
trips_2019_v2 %>% 
  group_by(usertype, sex) %>% 
  summarise(Rides = n()) %>% 
  arrange(usertype, desc(Rides)) %>% 
  ggplot(aes(x = sex, y = Rides, fill = usertype)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#50CAF7","#F59669","gray")) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Usertype by Sex, Number of Rides", subtitle = "2019", 
       x = " ", y = "Number of Rides", fill = " ")
```
<br>
<br>
Make a bar graph comparing sexes for casuals:
<br>
```{r}
trips_2019_v2_casuals %>% 
  group_by(usertype, sex) %>% 
  summarise(Rides = n()) %>% 
  arrange(usertype, desc(Rides)) %>% 
  ggplot(aes(x = sex, y = Rides, fill = usertype)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("#50CAF7","#F59669")) +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Casuals by Sex, Number of Rides", subtitle = "2019", 
       x = " ", y = "Number of Rides", fill = " ")
```

<br>
<br>
<br>
<br>